---
title: "Bivariate Viz - Exercises"
format: html
---

# Import data

```{r}
survey <- read.csv("https://hash-mac.github.io/ds112site-f26/data/survey.csv")
```

# How many students have filled out the survey?
44

# What type of variables do we have?
2 categorical 2 numerical

# Attach the package needed to use the ggplot function
```{r}
library(tidyverse)
```
#Example 1 - hangout preferences
```{r}
ggplot(survey, aes(x = hangout)) +
         geom_bar(color = "white", fill = "pink")
```

#Example 2 - temperature preferences
```{r}
#Plot 1
ggplot(survey, aes(x = fav_temp_c)) +
  geom_bar(color = "cyan")

#Plot 2
ggplot(survey, aes(x = fav_temp_c)) + 
  geom_histogram(color = "white", fill = "magenta")
```

#Exercise 0

```{r}
# Load data
elections <- read.csv("https://mac-stat.github.io/data/election_2020_county.csv")

# Check it out
head(elections)
```
#part a
- GUESS: 65%
- winner variable plot

```{r}
ggplot(elections, aes(x = winner_20)) +
  geom_bar(color = "white", fill = "red")
```
- 83%
#part b - repub_pct_20 - % republican vote per county in 2020

```{r}
ggplot(elections, aes(x = repub_pct_20)) +
  geom_histogram(color = "white", fill = "red")
```

#exercise 1 - quant vs quant intuition check
```{r}
ggplot(elections, aes(x = repub_pct_16, y = repub_pct_20)) +
  geom_point(color = "red")
```

#exercise 2 - 2 quant variables - repub_pct_20 vs. repub_pct_16 - different glyphs for each county

# Set up the plotting frame
# How does this differ from the frame for our histogram of repub_pct_20 alone?
```{r}
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16))
```

# Add a layer of points for each county
# Take note of the geom! - dots
```{r}
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point()
```
```{r}
# Change the shape of the points
# What happens if you change the shape to another number?
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point(shape = 8, color = "green")
```

```{r}
# Add a layer that represents each county by the state it's in
# Take note of the geom and the info it needs to run!
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_text(aes(label = state_abbr), color = "pink")
```

#exercise 3 - reflection
- strong relationship
- positive relationship
- texas, idaho, and utah went even more republican in 2020 but otherwise stayed consistent

#exercise 4 - visualizing trend
```{r}
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point() +
  geom_smooth()
```
#part a - new plot with the smooth line but not individual point glyphs

```{r}
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_smooth()
```

#part b - best linear model - method = "lm"

```{r}
ggplot(elections, aes(y = repub_pct_20, x = repub_pct_16)) +
  geom_point() +
  geom_smooth(method = "lm")
```
  

#exercise 5

```{r} 
#repub_pct_20 vs. median_rent
ggplot(elections, aes(x = repub_pct_20, y = median_rent)) +
  geom_point() +
  geom_smooth(method = "lm")

#repub_pct_20 vs. median_age
ggplot(elections, aes(x = repub_pct_20, y = median_age)) +
  geom_point() +
  geom_smooth(method = "lm")
```
Rent is a better predictor; negative relationship

#exercise 6 - sad scatterplot - repub_pct_20 (quant) vs. historical (categorical)

- two different classes --> scatterplot may not clearly represent change over time 

```{r}
ggplot(elections, aes(y = repub_pct_20, x = historical)) +
  geom_point()
```
-shows general skew but dots are too close together to gather anything beyond base-level analysis

#exercise 7 - quantitative vs. categorical - violins & boxes - 2 new geom layers

```{r}
# Side-by-side violin plots
ggplot(elections, aes(y = repub_pct_20, x = historical)) +
  geom_violin()
```
```{r}
# Side-by-side boxplots (defined below)
ggplot(elections, aes(y = repub_pct_20, x = historical)) +
  geom_boxplot()
```

- top of line: maximum
- top of box: 75th percentile
- line in box: median
- bottom of box: 25th percentile
- bottom of line: minimum
- dots outside: outlier(s)

- historically blue states' median vote was slightly over 50% republican in 2020, as were purple states (but moreso). historically red states' median vote across counties was around 75% republican. 

-the largest range/diversity of voting was in blue states

#exercise 8 - quant vs cat intuition check

```{r}
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density()
```
#exercise 9: quant vs. categorical - density plots

```{r}
# Name two "bad" things about this plot - 1. you can't see the values hidden by the fill 2. the y-axis doesn't provide a communicable scale/reference point - unclear what high density means here
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density()
```
```{r}
# What does scale_fill_manual do? - allows you to choose the colors manually
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density() +
  scale_fill_manual(values = c("blue", "purple", "red"))
```
```{r}
# What does alpha = 0.5 do? - changes opacity level! you can now see the hidden values beneath the other colors
# Play around with different values of alpha, between 0 and 1
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density(alpha = 0.1) +
  scale_fill_manual(values = c("blue", "purple", "red"))
```
```{r}
# What does facet_wrap do?! - separates the density plots as 3 separate graphics
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_density() +
  scale_fill_manual(values = c("blue", "purple", "red")) +
  facet_wrap(~historical)
```
```{r}
# Let's try a similar grouping strategy with a histogram instead of density plot.
# Why is this terrible? - you can't see the values underneath, and alpha doesn't help!!
ggplot(elections, aes(x = repub_pct_20, fill = historical)) +
  geom_histogram(color = "white", alpha = 0.1) +
  scale_fill_manual(values = c("blue", "purple", "red"))
```

#exercise 10 - reflect
- i like the density plot with opacity fill because it's visually pleasing, but box plots are more informative
- density plots show relative scale of each trend 
- box plots show the character of the data (its median, its range, its outliers)

#exercise 11 - cat vs cat
```{r}
# Plot 1: adjust this to recreate the top plot
ggplot(elections, aes(x = historical, fill = winner_20)) +
  geom_bar()
```
```{r}
# Plot 2: adjust this to recreate the bottom plot
ggplot(elections, aes(x = winner_20)) +
  geom_bar() +
  facet_wrap(~historical)
```

#exercise 12 - cat vs. cat

```{r}
# A stacked bar plot
# How are the "historical" and "winner_20" variables mapped to the plot, i.e. what roles do they play? - historical groups the results as historical blue/red/purple counties, and winner_20 shows the winner in those counties in 2020
ggplot(elections, aes(x = historical, fill = winner_20)) +
  geom_bar()
```

```{r}
# A faceted bar plot
ggplot(elections, aes(x = winner_20)) +
  geom_bar() +
  facet_wrap(~historical)
```
```{r}
# A side-by-side bar plot
# Note the new argument to geom_bar
ggplot(elections, aes(x = historical, fill = winner_20)) +
  geom_bar(position = "dodge")
```

```{r}
# A proportional bar plot
# Note the new argument to geom_bar
ggplot(elections, aes(x = historical, fill = winner_20)) +
  geom_bar(position = "fill")
```
#part a - pro of proportional bar plot: out of 100, so scale is clear; con: results might appear larger than they actually are because this is merely a proportion 

#part b - historical stacked bar - reflects true values relative to each other, separated by historically red/blue/purple counties

#exercise 13:
```{r}
weather <- read.csv("https://mac-stat.github.io/data/weather_3_locations.csv")
```

```{r}
# How do 3pm temperatures (temp3pm) differ by location? - num + cat = density 
ggplot(weather, aes(x = temp3pm, fill = location)) +
  geom_density(alpha = .1)
```

```{r}
# How might we predict the 3pm temperature (temp3pm) by the 9am temperature (temp9am)? - num vs. num -> scatter + line
ggplot(weather, aes(x = temp9am, y = temp3pm)) +
  geom_point() +
  geom_smooth(method = "lm")
```
```{r}
# How do the number of rainy days (raintoday) differ by location?
ggplot(weather, aes(y = raintoday, fill = location)) +
  geom_bar()
```
#exercise 14: state support
```{r}
ggplot(elections, aes(x = repub_pct_20, fill = state_abbr)) + 
  geom_density(alpha = 0.5)
```
--> too many states for clear viz

```{r}
ggplot(elections, aes(x = repub_pct_20)) + 
  geom_density(alpha = 0.5) + 
  facet_wrap(~ state_abbr)
```
--> not relative to each other; hard to analyze

```{r}
# Remember to install ggridges package before calling it
library(ggridges)

# Make our first joy plot (ridgeline - quantitative outcome among MANY groups + categories!)
# THINK: still hard to read because of how many states there are; scale on small R window looks ugly; reverse ABC order
ggplot(elections, aes(x = repub_pct_20, y = state_abbr)) + 
  geom_density_ridges()
```
```{r}
# Let's put the states in order by Republican support, not alphabet
# How do you think fct_reorder works? We'll learn about this later in the semester. - factors reordered based on its relation to another value
ggplot(elections, aes(x = repub_pct_20, y = fct_reorder(state_abbr, repub_pct_20))) + 
  geom_density_ridges(alpha = 0.5)
```

```{r, fig.width=8, fig.height=10}
# YOUR TURN: color/fill the ridges according to a state's historical voting patterns 
# and add meaningful axis labels
ggplot(elections, aes(x = repub_pct_20, y = fct_reorder(state_abbr, repub_pct_20), fill = historical)) + 
  geom_density_ridges(alpha = 0.5) + 
  xlab("Republican Vote % in 2020") +
  ylab("State") + 
  scale_fill_manual(values = c("blue", "purple", "red"))

```
- largest variability among counties: MA, NJ, UT, AL, MS, AZ
- least variability: HI, NH, TN, IN, WV
- can the states be grouped based on least variability?

#exercise 15: state total outcomes
```{r}
elections_by_state <- read.csv("https://mac-stat.github.io/data/election_2020_by_state.csv")
```

```{r}
ggplot(elections_by_state, aes(y = repub_pct_20, x = repub_pct_16)) + 
  geom_point()
```

```{r, fig.width=8, fig.height=10}
# YOU TRY
# Start by creating a "scatterplot" of state_abbr (y-axis) by 2020 Republican support on the x-axis
# Color the points red
# Scroll to solutions below when you're ready

ggplot(elections_by_state, aes(y = state_abbr, x = repub_pct_16)) +
  geom_point(color = "red")
```

```{r, fig.width=8, fig.height=10}
# YOU TRY
# Reorder the states in terms of their 2020 Republican support (not alphabet)
# Scroll to solutions below when you're ready
ggplot(elections_by_state, aes(x = repub_pct_20, y= fct_reorder(state_abbr, repub_pct_20))) + 
  geom_point(color = "red")
```

```{r, fig.width=8, fig.height=10}
# Finally, add ANOTHER layer of points for the 2016 outcomes
# What info does this new geom_point() layer need to run?
ggplot(elections_by_state, aes(x = repub_pct_20, y = fct_reorder(state_abbr, repub_pct_20))) + 
  geom_point(color = "red") + 
  geom_point(aes(x = repub_pct_16, y = state_abbr), shape = 3, color = "maroon") + 
  xlab("Republican Vote % in 2016 (maroon crosses) and 2020 (red dots)") +
  ylab("State")
```
- biggest change between 2016-2020: HI (more republican), ID (more republican), NM (more republican), CO (less republican)

- how can it be grouped by % change?
